HiC Contact Map Comaprison Using Graphlet Approach

### 2018-04-18 papers/manuscripts/manuscript.tex
modified manuscript

@@

\cite{milenkoviae2008uncovering} 
 provide heuristics to compare two nodes based on some feature
(or signature) vectors, which is a 73-dimensional vector
$\mathbf{s}^T
= [s_0, s_2, ..., s_{72}]$ where $s_i$ denotes the number of nodes in
the network that are part of an orbit $i$. \\
\textit{Important Result}: Proteins with similar surroundings perform
similar functions.

@@

In order to threshold the matrix so that both global and local patterns are
captured, we borrowed the concept of \textit{adaptive thresholding} from image 
processing context. In this method, in order to be set, a pixel should have
an intensity larger than the average of non-zero intensities in its
\textit{neighborhood}. The neighborhood is defined by a sliding kernel 
that passes through the contact map with the pixel at its middle at 
each step. Figure \ref{local_thresholded_chr1_chr1} demonstrates result of 
this thresholding approach for intra-chromosomal contact maps of chromosome 1.
Refer to supplementary material for all 23 interchromosomal thresholding
results.
\begin{figure}[t]
    \centering
    \includegraphics[width=.5\textwidth]{figures/local_thresholded_chr1_chr1.png}
    \caption{Result of thresholding interchromosomal contact map of chromosome 1
    using a kernels of size $5 \times 5$ for all cell lines. 
    The first row shows the thresholded
    maps. Second and third rows demonstrate pair-wise similarities and 
    differences between contact maps respectively.}
    \label{local_thresholded_chr1_chr1}
\end{figure}

@@

\subsection{Statistical Analysis}
A one sided paired t-test
was conducted in order to compare pair-wise distances to
zero and also to each other. 
The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. The results of tests for $C_{1,1}$ 
is shown in table \ref{tab:t-test}
Results of all t-tests for each
contact map can be found in supplementary materials.
For each contact map, each pair of cells are
compared based on whether they are statistially
larger that the other or not with a signifiance
of $0.01$.

\begin{table}[]
    \centering
    \resizebox{.5\textwidth}{!}{
    \begin{tabular}{l|ccccccc}
        \rule{0pt}{25pt}
                  & One       & MIT-RL       & ALL-RL       & ALL-MIT      & ALL-CALL4    & CALL4-MIT    & CALL4-RL     \\ \hline
        \rule{0pt}{25pt}
        One       & -         & \textgreater & \textgreater & \textgreater & \textgreater & \textgreater & \textgreater \\
        \rule{0pt}{25pt}
        MIT-RL    & \textless & -            & \textless    & \textless    & \textless    & \textless    & \textless    \\
        \rule{0pt}{25pt}
        ALL-RL    & \textless & \textgreater & -            & \textless    & \textless    & \textless    & \textless    \\
        \rule{0pt}{25pt}
        ALL-MIT   & \textless & \textgreater & \textgreater & -            & =            & =            & =            \\
        \rule{0pt}{25pt}
        ALL-CALL4 & \textless & \textgreater & \textgreater & =            & -            & =            & =            \\
        \rule{0pt}{25pt}
        CALL4-MIT & \textless & \textgreater & \textgreater & =            & =            & -            &              \\
        CALL4-RL  & \textless & \textgreater & =            & =            & =            & =            & -           
        \rule{0pt}{25pt}
    \end{tabular}}
    \caption{
        Result of one-sided t-test 
        with significance $\alpha = 0.01$
        for intrachromosomal
        contact map of chromosome 1.
        As an example interpretation,
        $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ 
        is statistically larger than 0,
        but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ . 
        An equal sign actually means that the two vectors are
        not statistically different from each other.
        We can also conclude that 
        $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
        statistically different from 
        $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
        between graphlets extracted from all contact maps. 
        Refer to supplementary
        material for results of all hypotheis tests.
    }
    \label{tab:t-test}
\end{table}
We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1  and whether the average 
correlations across orbits is equal to 1. The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the 
full list of t-test results.

>>>>

Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge is that 
3D structure of a cell is not readily available. Based on
\cite{adhikari2016chromosome3d}, fluorescence in situ hybridizaiton
(FISH) is used for investigating 3D configuration of chromosomes.
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.
In orther to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. Based on HiC data, an
\textit{interaction frequency (IF) } matrix can be developed between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both inter- and intra-chromosomal interaction matrices.
We believe differences in interaction matrices can be found between normal cells and cancerous ones.

In this study, we plan to find dissimilarities between normal cells and cancerous cells,
through investigating HiC contact maps. 
We suspect that there are systematic differences between how chromosomes are structured
between normal cells and cancerous cells.
we 
Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge that we face is that 
3D structure of a cell is not readily available. Based on
\cite{adhikari2016chromosome3d}, fluorescence in situ hybridizaiton
(FISH) is used for investigating 3D configuration of chromosomes.
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.
In orther to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by \cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. Based on HiC data, an
\textit{interaction frequency (IF) } matrix can be developed between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both inter- and intra-chromosomal interaction matrices.
We believe differences in interaction matrices can be found between normal cells and cancerous ones.

@@

\cite{milenkoviae2008uncovering} 
 provide heuristics to compare two nodes based on some feature
(or signature) vectors, which is a 73-dimensional vector
$\mathbf{s}^T
= [s_0, s_2, ..., s_{72}]$ where $s_i$ denotes the number of nodes in
the network that are part of an orbit $i$. \\
They concluded that proteins with similar surroundings perform
similar functions.

@@

To the best of our knowledge, little work has dealt with the task of
thresholding HiC contact maps. There has been some statistical approaches
developed on similar data in other fields. 
For example, authors of \cite{ginestet2011statistical} developed 
Statistical Network (SPN) analysis
where the choice of thresholding value is made by statistical inference.
This method, although very robust, works within the framework of 
design of experiments where the same
network can be extracted for different 
individuals under different treatments. Thus a relatively large
set of different contact maps need to be available in order for
this method to be applicable towards our end.

Instead, in order to threshold the matrix so that both global and local patterns are
captured, we borrowed the concept of \textit{adaptive thresholding} from image 
processing context. In this method, in order to be set, a pixel should have
an intensity larger than the average of non-zero intensities in its
\textit{neighborhood}. The neighborhood is defined by a sliding kernel 
that passes through the contact map with the pixel at its middle at 
each step. Figure \ref{local_thresholded_chr1_chr1} demonstrates result of 
this thresholding approach for intra-chromosomal contact maps of chromosome 1.
Refer to supplementary material for all 23 interchromosomal thresholding
results.
\begin{figure}[t]
    \centering
    \includegraphics[width=.5\textwidth]{figures/local_thresholded_chr1_chr1.png}
    \caption{Result of thresholding interchromosomal contact map of chromosome 1
    using a kernels of size $5 \times 5$ for all cell lines. 
    The first row shows the thresholded
    maps. Second and third rows demonstrate pair-wise similarities and 
    differences between contact maps respectively.}
    \label{local_thresholded_chr1_chr1}
\end{figure}

@@

\subsection{Statistical Analysis}
In order to quntify our results, we have conducted one-sided
t-test in order to test whether the average pair-wise
MIC values across orbits are significantly different from 
each other and also from 1. 
however, pairwise correlation of cancer cells for
orbits in $\mathbb{Q}$ were significantly higher than
correlation between cancer cells and normal cells.
The results
for this test showed that all values are significantly less than 1;
A summary of can be found in
table \ref{tab:t-test}.
We have also conducted t-tests for 
pair-wise orbit signature vector distances across contact maps.
Please refer to supplementary material for result of the 
full list of t-test results.
\begin{table}[H]
    \resizebox{.5\textwidth}{!}{
\begin{tabular}{c|ccccccc}
    & $MIC_{\scriptscriptstyle MIT-CALL4}$ & $MIC_{\scriptscriptstyle MIT-RL}$ & $MIC_{\scriptscriptstyle MIT-ALL}$ & $MIC_{\scriptscriptstyle ALL-CALL4}$ & $MIC_{\scriptscriptstyle CAll4-RL}$ & $MIC_{\scriptscriptstyle ALL-RL}$ & 1   \\
              \rule{0pt}{1pt} \\ \hline \rule{0pt}{15pt}
              $MIC_{\scriptscriptstyle MIT-CALL4}$ & -         & $\mathbf{<}$    & $\mathbf{<}$     & $\mathbf{<}$       & $\mathbf{<}$      & $\mathbf{<}$    & $\mathbf{<}$ \\
              \rule{0pt}{20pt}
              $MIC_{\scriptscriptstyle MIT-RL}$    & $\mathbf{>}$       &   -    & $\mathbf{<}$     & $\mathbf{<}$       & $\mathbf{<}$      & $\mathbf{<}$    & $\mathbf{<}$ \\
              \rule{0pt}{20pt}
              $MIC_{\scriptscriptstyle MIT-ALL}$   & $\mathbf{>}$       & $\mathbf{>}$    &    -    & $\mathbf{<}$       & $\mathbf{<}$      & $\mathbf{<}$    & $\mathbf{<}$ \\
              \rule{0pt}{20pt}
              $MIC_{\scriptscriptstyle ALL-CALL4}$ & $\mathbf{>}$       & $\mathbf{>}$    & $\mathbf{>}$     &       -   & $\mathbf{<}$      & $\mathbf{<}$    & $\mathbf{<}$ \\
              \rule{0pt}{20pt}
              $MIC_{\scriptscriptstyle CAll4-RL}$  & $\mathbf{>}$       & $\mathbf{>}$    & $\mathbf{>}$     & $\mathbf{>}$       &    -     & $\mathbf{<}$    & $\mathbf{<}$ \\
              \rule{0pt}{20pt}
              $MIC_{\scriptscriptstyle ALL-RL}$    & $\mathbf{>}$       & $\mathbf{>}$    & $\mathbf{>}$     & $\mathbf{>}$       & $\mathbf{>}$      &   -    & $\mathbf{<}$ \\
              \rule{0pt}{20pt}
              1         & $\mathbf{>}$       & $\mathbf{>}$    & $\mathbf{>}$     & $\mathbf{>}$       & $\mathbf{>}$      & $\mathbf{>}$    &    -
\end{tabular}}
    \caption{
        Result of one-sided t-test 
        comparison of MIC results 
        with significance $\alpha = 0.01$
        for orbits of $\mathbb{Q}$.
        A $<$ sign means that the value in the corresponding
        row is significantly smaller that the value in 
        corresponding column; and a $>$ sign means the opposite.
        An $=$ sign would have meant the values are not 
        statistically different.
        As can be seen, all MIC values are significantly smaller that 1,
        meaning non on them are exactly the same; 
        however, all Cancer vs Cancer MIC values are statistically
        larger than Cancer vs Normal, implying that 
        spatial positioning of loci are more similar between cancer cells.
    }
    \label{tab:t-test}

\end{table}
